PhD Proposal

### 2022-04-20 abstract/abstract.tex
continuo abstract

\documentclass[a4paper,12pt, oneside]{article}
% \usepackage{fullpage}
%\usepackage[italian]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE,RO]{\slshape \rightmark}
\fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PHD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}

@@

\section{Introduction}

@@

\textbf{run-length version of BWT (\textit{RLPBWT})} with the ultimate intention

@@

a text. A recent improvement, regarding the \textit{RLPBWT}, has been made

@@

\section{State of the art}

@@

\subsection*{RLBWT, Matching Statistics, MONI and PHONI}

@@

\subsection*{RLPBWT}
\section{Research goals}
% RLMPBWT
% missing data

>>>>

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=20mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE,RO]{\slshape \rightmark}
\fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}

@@

\section*{Introduction}

@@

\textbf{run-length version of BWT (\textit{RLBWT})} with the ultimate intention

@@

a text. A recent improvement, regarding the \textit{RLBWT}, has been made

@@

\section*{State of the art}

@@

\subsection*{RLBWT}
Speaking of pangenome, linear indexing via FM-index is no longer the best
solution as it does not handle the large repetitions present in this new type of
sequences in the best possible way. In 2005 M\"{a}niken and Navarro defined
the \textbf{Run-Length Burrows–Wheeler Transform (\textit{RLBWT})}. Given a text
$T$, $RLBWT_T$ is a rappresentation of $BWT_T$ with a compact rappresentation of
consecutive equal characters, the so-called \textit{runs}. With this change of
perspective the algorithms have gone from being linear over the length of the
text, $n$, to being linear over the number of runs, $r$. \\
The new indexing method, which was then introduced by Gagie et
al. \cite{rindex}, was called \textbf{r-index} and corresponds to the
\textit{RLBWT} and a SA sampling at the begin and at the end of every run. The
algorithm for querying through the \textit{RLBWT} take advantage of other
methods, such as the use of \textbf{thresholds} (minimum
\textit{LCP} value between two consecutive runs of the same character) in
\textit{MONI}, and the use of \textbf{longest common extension (\textit{LCE})
  query} (to compute the right equal common extension between two position in
the text) cd\cite{lce} and \textbf{straight-line programs (\textit{SLP})}
\cite{slp} (for \textit{random access} and \textit{lce queries} over the
grammar-compressed text) in \textit{PHONI}. Both these solutions are used for
the computation of the \textbf{matching statistics (\textit{MS})}. Given a
pattern $P$ and a text $T$ the \textit{MS} of $P$ in respect to $P$ is an array
$M$ of pairs position/length, $|M|=|P|$, such that
$T[M[i].pos:M[i].pos+M[i].len-1=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$ does
not occur in $T$. Given \textit{MS} we can compute every \textbf{Maximal Exact
  Match (\textit{MEM})} of a pattern in a text. Given a text $T$ and a pattern
$P$ a substring of the pattern $P[i : i+l-1]$, of length $l$, is a \textit{MEM}
of $P$ in $T$ if $P[i:i+l-1]$ is a substring of $T$ but but neither
$P[i-1:i+l-1]$ nor $P[i:i+l]$ are. At the end, using a particular function
called $\varphi$ (and $\varphi^{-1}$), based on the use of the \textbf{inverse
  suffix array (\textit{ISA})} as well, it was possible to start from the
starting positon of a match extracted 
by \textit{MS} and find all the starting positions of all the copies of $P$ in
$T$. Infact, formally, given a starting position $p$, we have
$\varphi(p)=SA[ISA[p]-1]$, $NULL$ if $ISA[p]=0$, and
$\varphi^{-1}(p)=SA[ISA[p]+1]$, $NULL$ if $ISA[p]=|T|-1$, where $ISA[i]=j$ iff
$SA[j]=i$.\\   
Thanks to these and other methods it was possible to perform pattern matching
efficiently even on long sequences of nucleotides, such as those studied in
a pangenomic context. 

@@

Based on the theories of BWT Durbin, in 2014, he devised the positional
\textbf{Burrows–Wheeler transform (\textit{PBWT}}) \cite{pbwt}, in order to
solve the problem of pattern matching on panels (matrices) of haplotypes,
denoted by $X$. In
detail we are talking about a panel with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a reversed-prefix ordering at each column
$k$ that produces two different multidimensional arrays. The first one is called
\textbf{prefix array}, denoted by $a$, and, for each column $k$, contains, for
each position $i$, the haplotype of index $m$ in the original panel. More
formally we can say that $a_k[i]=m$ iff $X_m$ is the $i$-th haplotype in the
reversed-prefix ordering at column $k$. Note that $x_m$ such that $a_k[i]=m$
could be denoted by $y_i^k$. The second bidimensional array is
called \textbf{divergence array}, denoted by $d$, and indicates the index of the
starting column of the longest common suffix, ending in column $k$, between a
row and its previous one, at reversed-prefix ordering at column $k$. More
formally we can define $d_k[i]=h$ iff $h$ is the smallest column index such that
$y_i^k[h,k)=y_{i-1}^k[h,k)$. (\textbf{RISCRIVERE MEGLIO})\\
Thanks to these two bidimensional arrays it is possible to compute all matches
within $X$ longer than a minimum length $L$, all set-maximal matches within $X$
in linear time, all set-maximal matches (which we could also call ``MEMs'') from
a new sequence $z$ to $X$ etc$\ldots$ 
\section*{Research goals}
For my master's thesis  I therefore tried to combine the ideas related to the
\textit{RLBWT} with those of the \textit{PBWT}, creating the \textbf{RLPBWT}. In
order to obtain this result I have to rethink the concept of \textit{Matching
  Statistics} for \textit{PBWT}, how to compute the \textit{SLP} for the panel,
how to use \textit{thresholds}/\textit{LCE queries}, how to obtain the same
behaviour of the $\varphi$ function etc$\ldots$\\
In detail, regarding \textit{MS}, instead of the \textit{pos} we track a
\textit{row} of the panel, regarding thresholds we use the index inside a run in
a column where we have the minimum value of divergence array and, regarding the
\textit{SLP}, we ``stretch'' the reverse panel (from the right to the left) in
in order to make \textit{LCE queries} possible. Instead, regarding the $\varphi$
function, I have implemented a new simple data structure to identify which row
is above and which row is below each row in the panel permutated via the
prefix array and stored in a run-length compressed way. (\textbf{RISCRIVERE
  MEGLIO ED EVENTUALMENTE APPROFONDIRE}).\\
Obviously there are some limitations, such as the study of biallelic panels
only and the lack of management of any missing data, which are very frequent in
the case of real data. First of all it will be interesting to implement a new
version of the \textit{RLPBWT} that can handle multiallelic data, adapting the
current use of bitvectors to manage the \textit{LF-mapping} also in that
case. On the other hand, it will be more complicated to manage the missing
data. This problem is known to be \textit{hard} so I should probably deal with
parametric algorithms or approximate algorithms, based on researches already
developed in \textit{BIAS}.
